Analysis of $\piup$-$\piup$ lattice QCD \\ scattering data

### 2015-03-17 Paper.tex
Restructure Lüscher formula section

\subject{physics760 Computational Physics}
\title{Analysis of $\piup$-$\piup$ lattice QCD scattering data}
%\subtitle{}
\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
}

@@

In order to get rid of all those problems at the same time one does a meta
analysis of the data itself. To get back to the example of the mean, let $X$ be
the set of $N$ data points ($x_i \in \R$). Then the estimator of the mean
$\bar\mu$ can be computed directly from $X$. Now $R$ bootstrap samples $X^r$
(upper index, no power) are computed from the set by randomly selecting $N$
elements from $X$ and putting them into $X^r$. Since $X$ is sampled from the
original distribution, it resembles the original distribution as good as it can
with only $N$ elements. $X^r$ therefore is sampled from an approximation of the
original distribution. Then the estimators for the means $\bar\mu^r$ are
computed from the bootstrapped sets. The mean of all the $\bar\mu^r$ should
give $\bar\mu$, which was computed from the original data. The error estimate
is then the standard deviation of all the $\bar\mu^r$.

@@

For the correlated fit, a new likelihood function is needed. I chose to keep
the least squared likelihood function but incorporate the correlation into the
$\chi^2$. First, a correlation matrix $\tens C$ is needed, which is computed
from the $R$ bootstrap samples:
\begin{equation}
    \label{eq:correlation_matrix}
    C_{ij} := \frac{1}{R[R-1]} \sum_{r=1}^R
    [x_{ir} - \bar x_{iR}] [x_{jr} - \bar x_{jR}]
    \eqnsep
    \bar x_{iR} := \frac 1R \sum_{r=1}^R x_{ir}.
\end{equation}

@@

Equation~\eqref{eq:luescher} can be motivated like so: Let $\hat H = \hat H_0 +
\hat V$ be the Hamiltonian for two identical particles where $\hat H_0$ is the
free Hamiltonian and $\hat V$ a spherically symmetric potential. The
eigenfunctions of $\hat H_0$ are plane waves which are denoted $\ket{\vec p,
\vec q}$. In continous space, the normalization is
\parencite[(2.10)]{luescher/volume_dependence}
\[
    \braket{\vec p', \vec q'|\vec p, \vec q}
    = [2\piup]^6 \sbr{
        \delta(\vec p' - \vec p)
        \delta(\vec q' - \vec q)
        +
        \delta(\vec p' - \vec q)
        \delta(\vec q' - \vec p)
    }.
\]
On the discrete lattice, the possible momentum values are restricted to lattice
values and the potential will be periodic due to periodic boundary conditions.
This changes the normalization to
\parencite[(2.21)]{luescher/volume_dependence}
\[
    \braket{\vec p', \vec q'|\vec p, \vec q}
    = L^6 \sbr{
        \delta_{\vec p', \vec p}
        \delta_{\vec q', \vec q}
        +
        \delta_{\vec p', \vec q}
        \delta_{\vec q', \vec p}
    }.
\]

The matrix element of the potential $\vec V$ are given by
\parencite[(2.21)]{luescher/volume_dependence}
\[
    \braket{\vec p', \vec q'| \hat V | \vec p, \vec q}
    = L^3 \delta_{\vec P', \vec P} \hat V\del{\frac12 [\vec p' - \vec q'],
    \frac12 [\vec p - \vec q]},
\]
where $\vec P$ is the total momentum of the two particles. The energy shift in
first order of $V$ for the ground state (all momenta are zero) then is the
quotient \parencite[(2.24)]{luescher/volume_dependence}:
\[
    \Deltaup E = \frac{1}{2L^3} \hat V(\vec 0, \vec 0) + \mathcal O(V^2).
\]
The scattering amplitude as well as the S-wave scattering length $a_0$ can be
expanded in powers of $\hat V$ (Born series). Using first order term from 
\parencite[(2.18)]{luescher/volume_dependence}, above equation can be rewritten
with the scattering length
\parencite[(2.25)]{luescher/volume_dependence}:
\[
    \Deltaup E = - \frac{4\piup a_0}{m L^3} + \mathcal O(V^2).
\]

To first order on $V$, the leading term scales with $L^{-3}$, like in
Equation~\eqref{eq:luescher}. To obtain more terms, a complete perturbation
expansion needs to be written out. That expansion will be both in terms of
$L^{-1}$ \emph{and} $V$. \citeauthor{luescher/volume_dependence} then was able
to group those terms in powers of $L^{-1}$ and obtain
Equation~\eqref{eq:luescher}.

% FIXME Maybe write a bit more on the methods?

@@

The results from Table~\ref{tab:computed} are shown in Figure~\ref{fig:result}.

>>>>

\subject{physics760 Computational Physics}
\title{Analysis of $\piup$-$\piup$ lattice QCD \\ scattering data}
%\subtitle{}
\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
}

@@

In order to get rid of all those problems at the same time one does a meta
analysis of the data itself. To get back to the example of the mean, let $X$ be
the set of $N$ data points ($x_i \in \R$). Then the estimator of the mean
$\bar\mu$ can be computed directly from $X$. Now $R$ bootstrap samples $X^r$
(upper index, no power) are computed from the set by randomly selecting $N$
elements from $X$ and putting them into $X^r$. Since $X$ is sampled from the
original distribution, it resembles the original distribution as good as it can
with only $N$ elements. $X^r$ therefore is sampled from an approximation of the
original distribution. Then the estimators for the means $\bar\mu^r$ are
computed from the bootstrapped sets. The mean of all the $\bar\mu^r$ should
give $\bar\mu$, which was computed from the original data. The error estimate
is then the standard deviation of all the $\bar\mu^r$.
\parencite[Slide~5]{Oser/Bootstrap}

@@

For the correlated fit, a new likelihood function is needed. I chose to keep
the least squared likelihood function but incorporate the correlation into the
$\chi^2$. First, a correlation matrix $\tens C$ is needed, which is computed
from the $R$ bootstrap samples: \parencite[Section~2]{Michael:1994sz}
\begin{equation}
    \label{eq:correlation_matrix}
    C_{ij} := \frac{1}{R[R-1]} \sum_{r=1}^R
    [x_{ir} - \bar x_{iR}] [x_{jr} - \bar x_{jR}]
    \eqnsep
    \bar x_{iR} := \frac 1R \sum_{r=1}^R x_{ir}.
\end{equation}

@@

Equation~\eqref{eq:luescher} can be motivated like so: Let $\hat H = \hat H_0 +
\hat V$ be the Hamiltonian for two identical particles where $\hat H_0$ is the
free Hamiltonian and $\hat V$ a spherically symmetric potential with a limited
range (or faster than $1/\sqrt r$ decay). Then the probability that the two
particles are in interaction range scales like $L^{-3}$ as this is the inverse
volume which the two particles occupy. Then the most prominent finite size
effect scales to the third power, all other effects can only have higher
powers.

In the first order of the potential $V$ the ground state (all momenta are zero)
will have an energy shift proportional to $L^{-3}$
\parencite[(2.24)]{luescher/volume_dependence}:
\[
    \Deltaup E = \frac{1}{2L^3} \hat V(\vec 0, \vec 0) + \mathcal O(V^2).
\]
The scattering amplitude $T$ as well as the S-wave scattering length $a_0$ can
be expanded in powers of $\hat V$ (Born series). Using first order term from
\parencite[(2.18)]{luescher/volume_dependence}, above equation can be rewritten
with the scattering length
\parencite[(2.25)]{luescher/volume_dependence}:
\[
    \Deltaup E = - \frac{4\piup a_0}{m L^3} + \mathcal O(V^2).
\]

To first order on $V$, the leading term scales with $L^{-3}$, like in
Equation~\eqref{eq:luescher}. To obtain more terms, a complete perturbation
expansion needs to be written out. That expansion will be both in terms of
$L^{-1}$ and $V$. \citeauthor{luescher/volume_dependence} then was able to
group those terms in powers of $L^{-1}$ and obtain
Equation~\eqref{eq:luescher}.

@@

The results from Table~\ref{tab:computed} are shown in Figure~\ref{fig:result}.
My data points have no marker, the reference values have a diamond marker. The
solid black line is the expectation, not a fit.

As can be seen with the comparison to the reference data, that my values are a
lot less accurate than the reference ones. For most points, my errors are a lot
larger. The statistical uncertainty on my values is high enough for them to be
compatible with the reference and expectation.
